Complementing theory on adolescent emotion regulation using  machine learning

### 2022-03-05 manuscript/manuscript.Rmd
1st comments
First (minor) comments to the manuscript.

Emotion regulation,
defined as the ability to manage emotional experience and expression,
is critical for mental health [@aldaoEmotionregulationStrategiesPsychopathology2010, @schaferEmotionRegulationStrategies2017, @braetEmotionRegulationChildren2014]
and social competence [@zimmermannEmotionRegulationEarly2014] throughout life.
There is mounting evidence, however, that adolescents do not uniformly become better at regulating emotions: 
staggered development of emotional and regulatory brain circuits gives rise to a "maturity gap" in mid-adolescence [@croneUnderstandingAdolescencePeriod2012].
The emotional salience of social stimuli is intensified, while the ability to regulate these emotions lags behind [@caseyBrakingAcceleratingAdolescent2011, @vanlissaLongitudinalInterplayAffective2014].
This imbalance motivates most adolescents to engage in adaptive exploration:
Seeking out experiences required to develop mature socio-emotional competences.
At the same time, this imbalance renders some adolescents vulnerable to emotion dysregulation [@leeAdolescentMentalHealth2014],
<!-- exposes youth to a risk of emotion dysregulation, -->
for example, because they are not yet well-equipped to regulate the pursuit of unhealthy incentives,
or persevere in pursuing adaptive goals when facing obstacles.
Thus, some become trapped in a negative spiral characterized by more negative and volatile emotions [@craccoEmotionRegulationChildhood2017],
which has been linked to the emergence of several mental disorders [@schaferEmotionRegulationStrategies2017].
<!-- This imbalance motivates most youngsters to seek out new challenges in life and love, -->
<!-- but renders some of them vulnerable to emotion dysregulation1,10. Unfortunately, there is little theory about which risk factors and environmental hazards render adolescents susceptible to emotional difficulties11. -->
<!-- Acquiring adaptive emotion regulation skills in adolescence plays a key role in this regard7,20, because t -->

@@

<!-- Maybe review theory briefly? -->
The first step towards theory formation or revision is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020].
<!--Thus, offering a starting point for this endeavor,<!-- empirical investigations of predictors of adolescents' emotion regulation development -->
A recent literature review did so
using a text mining systematic review of author keywords and abstracts of 6305 publications on  adolescents’ emotion regulation [@vanlissaMappingPhenomenaRelevant2021].
Typically, phenomena are identified using narrative literature reviews.
However, these are often limited to a small number of papers and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008]
Text mining systematic reviews, by comparison, are relatively comprehensive and derive insights from the literature through a relatively transparent and objective - albeit less interpretative - procedure [@vanlissaMappingPhenomenaRelevant2021]. <!-- reflection and meaning-making-->

@@

According to models of knowledge production within the quantitative positivist paradigm,
a field progresses through the cyclical alternation of exploratory and confirmatory research, also known as the "empirical cycle" [@degrootMethodologieGrondslagenVan1961, @wagenmakersCreativityVerificationCyclePsychological2018].
Exploratory research gives rise to testable predictions,
and confirmatory research evaluates testable predictions on new data.
Crucially, these phases are linked through evaluation of existing theory in light of empirical findings [@degrootMethodologieGrondslagenVan1961].
The fact that our field still lacks overarching theory despite decades of productive empirical research
suggests that the empirical cycle is not routinely completed.
In support of this claim, it has been noted that the overwhelming majority of research in our field appears, at first glance, to be confirmatory [@demuthQualitativeMethodologyDevelopmental2015].
More exploratory research is thus needed to complete the cycle.
Secondly, and paradoxically given the first argument,
the link between confirmatory research and theory is often tenuous.
For instance, many studies derive their predictions from prior empirical research, not from theory [e.g., @vanlissaCostEmpathyParentadolescent2017].
Thirdly, theories are rarely falsified or revised in light of empirical findings.
This failure to evaluate theory leads to flexible, "invulnerable" frameworks that can accommodate any result [@szollosiArrestedTheoryDevelopment2021].
To advance theory formation, our field might benefit from "closing the empirical cycle" through increased exploratory research and reflection on theory.

@@

From wave 2-6, adolescents completed the difficulty in emotion regulation scale [DERS, @gratzMultidimensionalAssessmentEmotion2004].
Based on prior psychometric analyses [see @vanlissaRoleFathersMothers2018],
we used 24 items of this 32-item scale
(e.g., "When I’m upset, I have difficulty thinking
about anything else").
Responses ranged
from 1 (Almost never) to 5 (Almost always).
Reliability was excellent in all waves, $\alpha \in [`r report(min(alphs), equals = F)`, `r report(max(alphs), equals = F)`]$.
We used latent variable scores (instead of mean scores) for further analyses.
To this end, we imposed a five-factor CFA (one for each wave) on the indicators of the dependent variable.
Latent means were restricted to zero, such that resulting scores reflect the deviation of individual adolescents' trajectories from the overall trajectory.
As measurement invariance must be imposed in order to compare latent scores across waves,
we compared a model with configural invariance ($BIC = 141247.46$)
to a model that imposed metric invariance ($BIC = 140913.27$).
The lower BIC indicates support for metric invariance.
The RMSEA and SRMR of the invariant model indicated acceptable fit ($\text{RMSEA} = .05, \text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
As reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
we considered it acceptable to proceed with individual factor scores for further analysis.

@@

The SEM-forests algorithm requires complete data and cannot accommodate multiple imputed data.
Consequently, we used single imputation using the missForest algorithm,
a random forests based approach with comparable performance to multiple imputation and FIML estimation [@stekhovenMissForestNonparametricMissing2012].
This method is valid when missingness is either random or contingent on observed data (MAR or MCAR).
It does not make any distributional assumptions and does not extrapolate,
meaning it returns realistic values within the range of the observed data.
The percentage of missingness ranged from $`r report(min((495-desc_cat[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for categorical data (see Table \@ref(tab:tabcat)),
and from $`r report(min((495-desc_num[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for continuous data (see Table \@ref(tab:tabnum)).
<!-- Original publication had up to 30% missing, Wolputte, E. V., & Blockeel, H. (2020, October). Missing value imputation with MERCS: a faster alternative to MissForest. In International Conference on Discovery Science (pp. 502-516). Springer, Cham. looked at higher numbers. According to Mplus: minimum 10% -->

@@

As we set out to identifying important predictors of emotion regulation development,
this latent class analysis is not sufficient to answer our research question.
In latent class analysis, individuals are clustered only with respect the the outcome (growth trajectories), to ensure maximum homogeneity of growth trajectories in each class.
It thus answers the question: "Which adolescents have similar growth trajectories?"
SEM-forests similarly aim to ensure maximum homogeneity of growth trajectories within clusters,
but they differ from latent class analysis in that they cluster individuals by reference to a set of predictors.
They thus answer the question: "Why do some adolescents have similar growth trajectories"?
This method is suitable to meet the study goals.

@@

There is prior evidence for the efficacy of conflict resolution training in adolescents [@breunlinConflictResolutionTraining2002].
The same evidence indicates that conflict resolution training caused a decrease in externalizing problems (another important predictor), suggesting potential compounding benefits.
Parenting behavior is another promising target for intervention,
as indeed it has historically been.
Importantly, our results suggest that such interventions should focus on promoting parental autonomy support, and decreasing intrusive, negative, guilt-inducing and overbearing parenting behaviors.
By contrast, promoting positive parenting practices may have limited potential for impact.
With this in mind, it is important to consider that positive parenting practices are often a key focus of parenting training.
Although one meta-analysis did report large effects of the Triple P-Positive Parenting Program on parental outcomes - including some negative parenting behaviors [@fletcherImpactBehaviouralParent2011],
another meta-analysis called these findings into question,
citing mixed evidence, a lack of adequate blinding, risk of bias, and the substantial cost of such programs [@wilsonHowEvidencebasedEvidencebased2012].
<!-- Another study Even when negative parenting has been assessed as outcomes, -->
<!-- significant effects were found for positive practices only [@marcynyszynGettingEvidencebasedProgram2011]. -->
<!-- and that there have been prior concerns about the efficacy of such training [@wilsonHowEvidencebasedEvidencebased2012; @cottamDiscoursesUnderpinningParenting2014]. -->
Also note that the efficacy of parenting training has been evaluated primarily in terms of parents' self-reported outcomes;
the ultimate benefit for children,
in particular for adolescents' emotion regulation,
is less studied.
The relatively higher importance rankings for negative versus parenting is in line with the well-established *negativity bias*,
that negative stimuli are more psychologically impactful than positive ones,
which has also been found in children [@vaishNotAllEmotions2008].
Future parenting interventions might take this into consideration by devoting sufficient attention to the elimination of negative parenting practices,
and including children's outcomes, including emotion regulation, into consideration.

@@

The present study had some unique strengths, most notably that it constituted the first comprehensive study of potential predictors of adolescent emotion regulation development, where prior work has taken more piecewise approaches, considering only a few predictors at a time.
<!-- The present study set out to identify the most important early indicators of developmental trajectories -->
<!-- of emotion regulation across adolescence. -->
<!-- This involves three important challenges: -->
<!-- First, the number of potentially relevant risk factors, as identified in a prior text mining systematic review, is large. -->
<!-- Second, there is a paucity of relevant theory to narrow down the candidate predictors. -->
<!-- Third, methods commonly used in the field are not well suited for large numbers of predictors. -->
<!-- Machine learning methods that perform variable selection can overcome these challenges. -->
<!-- We therefore used the SEM-forests algorithm to identify the most important early predictors of trajectories of emotion regulation development throughout adolescence. -->
<!-- The primary aim is to identify the most important predictors of adolescents' trajectories of emotion regulation development. -->
<!-- The secondary aim is to probe the nature of the marginal association of the predictors with these trajectories. -->

The study also has some limitations, most notably that the sample was selective,
which may limit the generalizability of the findings.
Specifically, this was a non-random, ethnically and socially relatively homogenous  sample from the Netherlands.
This raises some specific concerns, for example, that the low importance of SES might be due to a restriction of range.
The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world.
The only way to address these concerns is to replicate these findings in other samples.
Note that all code is available in a reproducible archive, which lowers the threshold to replication.
Another limitation is that the data source was a panel study with mostly questionnaire measures and relatively few biological, cognitive and neurological assays.
One final limitation related to the data is the relatively small sample size.
Machine learning methods benefit from large samples.
Although random forests are relatively robust to overfitting,
larger samples will likely yield more generalizable results,
and also afford more complicated models (thus removing the need for assumptions such as homogeneity of residuals).
Other limitations pertain to the method; for instance, SEM-forests require vast computational resources.
Computational constraints thus prevented us from tuning the model beyond default settings,
and from exploring marginal effects for higher order interactions.
Moreover, at the time of writing, there is not yet a method to evaluate the overall fit of a SEM-forest.

>>>>

Emotion regulation,
defined as the ability to manage emotional experience and expression,
is critical for mental health [@aldaoEmotionregulationStrategiesPsychopathology2010, @schaferEmotionRegulationStrategies2017, @braetEmotionRegulationChildren2014]
and social competence [@zimmermannEmotionRegulationEarly2014] throughout life.
There is mounting evidence, however, that adolescents do not uniformly become better at regulating emotions: 
staggered development of emotional and regulatory brain circuits gives rise to a "maturity gap" in mid-adolescence [@croneUnderstandingAdolescencePeriod2012].
The emotional salience of social stimuli is intensified, while the ability to regulate these emotions lags behind [@caseyBrakingAcceleratingAdolescent2011, @vanlissaLongitudinalInterplayAffective2014].
This imbalance motivates most adolescents to engage in adaptive exploration:
Seeking out experiences required to develop mature socio-emotional competences.
At the same time, this imbalance renders some adolescents vulnerable to emotion dysregulation [@leeAdolescentMentalHealth2014].
<!-- exposes youth to a risk of emotion dysregulation, -->
For example, they might not yet be sufficiently equipped to regulate the pursuit of unhealthy incentives,
or persevere in pursuing adaptive goals when facing obstacles.
Thus, some become trapped in a negative spiral characterized by more negative and volatile emotions [@craccoEmotionRegulationChildhood2017],
which has been linked to the emergence of several mental disorders [@schaferEmotionRegulationStrategies2017].
<!-- This imbalance motivates most youngsters to seek out new challenges in life and love, -->
<!-- but renders some of them vulnerable to emotion dysregulation1,10. Unfortunately, there is little theory about which risk factors and environmental hazards render adolescents susceptible to emotional difficulties11. -->
<!-- Acquiring adaptive emotion regulation skills in adolescence plays a key role in this regard7,20, because t -->

@@

<!-- Maybe review theory briefly? -->
The first step towards theory formation or revision is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020].
<!--Thus, offering a starting point for this endeavor,<!-- empirical investigations of predictors of adolescents' emotion regulation development -->
A recent literature review did so
using a text mining systematic review of author keywords and abstracts of 6305 publications on  adolescents’ emotion regulation [@vanlissaMappingPhenomenaRelevant2021].
Typically, phenomena are identified using narrative literature reviews.
However, these are often limited to a small number of papers and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008].
Text mining systematic reviews, by comparison, are relatively comprehensive and derive insights from the literature through a relatively transparent and objective - albeit less interpretative - procedure [@vanlissaMappingPhenomenaRelevant2021]. <!-- reflection and meaning-making-->

@@

According to models of knowledge production within the quantitative positivist paradigm,
a field progresses through the cyclical alternation of exploratory and confirmatory research, also known as the "empirical cycle" [@degrootMethodologieGrondslagenVan1961, @wagenmakersCreativityVerificationCyclePsychological2018].
Exploratory research gives rise to testable predictions,
and confirmatory research evaluates testable predictions on new data.
Crucially, these phases are linked through evaluation of existing theory in light of empirical findings [@degrootMethodologieGrondslagenVan1961].
The fact that our field still lacks overarching theory despite decades of productive empirical research
suggests that the empirical cycle is not routinely completed.
In support of this claim, it has been noted that the overwhelming majority of research in our field appears, at first glance, to be confirmatory [@demuthQualitativeMethodologyDevelopmental2015].
More exploratory research is thus needed to not only complete the cycle, but to enable confirmatory prediction testing growing from exploratory approaches.
Secondly, and paradoxically given the first argument,
the link between confirmatory research and theory is often tenuous.
For instance, many studies derive their predictions from prior empirical research, not from theory [e.g., @vanlissaCostEmpathyParentadolescent2017].
Thirdly, theories are rarely falsified or revised in light of empirical findings.
This failure to evaluate theory leads to flexible, "invulnerable" frameworks that can accommodate any result [@szollosiArrestedTheoryDevelopment2021].
To advance theory formation, our field might benefit from "closing the empirical cycle" through increased exploratory research and reflection on theory.

@@

From wave 2-6, adolescents completed the difficulty in emotion regulation scale [DERS, @gratzMultidimensionalAssessmentEmotion2004].
Based on prior psychometric analyses [see @vanlissaRoleFathersMothers2018],
we used 24 items of this 32-item scale
(e.g., "When I’m upset, I have difficulty thinking
about anything else").
Responses ranged
from 1 (Almost never) to 5 (Almost always).
Score reliability was excellent in all waves, $\alpha \in [`r report(min(alphs), equals = F)`, `r report(max(alphs), equals = F)`]$.
We used latent variable scores (instead of mean scores) for further analyses.
To this end, we imposed a five-factor CFA (one for each wave) on the indicators of the dependent variable.
Latent means were restricted to zero, such that resulting scores reflect the deviation of individual adolescents' trajectories from the overall trajectory.
As measurement invariance must be imposed in order to compare latent scores across waves,
we compared a model with configural invariance ($BIC = 141247.46$)
to a model that imposed metric invariance ($BIC = 140913.27$).
The lower BIC indicates support for metric invariance.
The RMSEA and SRMR of the invariant model indicated acceptable fit ($\text{RMSEA} = .05, \text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
As score reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
we considered it acceptable to proceed with individual factor scores for further analysis.

@@

The SEM-forests algorithm requires complete data and cannot accommodate multiple imputed data.
Consequently, we used single imputation using the missForest algorithm,
a random forests based approach with comparable performance to multiple imputation and FIML estimation [@stekhovenMissForestNonparametricMissing2012].
This method is valid when missingness is either random or contingent on observed data (MAR or MCAR).
It does not make any distributional assumptions and does not extrapolate,
meaning it returns realistic values within the range of the observed data.
The proportion of missingness ranged from $`r report(min((495-desc_cat[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for categorical data (see Table \@ref(tab:tabcat)),
and from $`r report(min((495-desc_num[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for continuous data (see Table \@ref(tab:tabnum)).
<!-- Original publication had up to 30% missing, Wolputte, E. V., & Blockeel, H. (2020, October). Missing value imputation with MERCS: a faster alternative to MissForest. In International Conference on Discovery Science (pp. 502-516). Springer, Cham. looked at higher numbers. According to Mplus: minimum 10% -->

@@

As we set out to identifying important predictors of emotion regulation development,
this latent class analysis is not sufficient to answer our research question.
In latent class analysis, individuals are clustered only with respect the the outcome (growth trajectories), to ensure maximum homogeneity of growth trajectories in each class.
It thus answers the question: "Which adolescents have similar growth trajectories?"
SEM-forests similarly aim to ensure maximum homogeneity of growth trajectories within clusters,
but they differ from latent class analysis in that they cluster individuals by reference to a set of predictors.
They thus answer the question: "Why do some adolescents have similar growth trajectories?"
This method is suitable to meet the study goals.

@@

There is prior evidence for the efficacy of conflict resolution training in adolescents [@breunlinConflictResolutionTraining2002].
The same evidence indicates that conflict resolution training caused a decrease in externalizing problems (another important predictor), suggesting potential compounding benefits.
Parenting behavior is another promising target for intervention,
as indeed it has historically been.
Importantly, our results suggest that such interventions should focus on promoting parental autonomy support, and decreasing intrusive, negative, guilt-inducing and overbearing parenting behaviors.
By contrast, promoting positive parenting practices may have limited potential for impact or even been detrimental.
With this in mind, it is important to consider that positive parenting practices are often a key focus of parenting training.
Although one meta-analysis did report large effects of the Triple P-Positive Parenting Program on parental outcomes - including some negative parenting behaviors [@fletcherImpactBehaviouralParent2011],
another meta-analysis called these findings into question,
citing mixed evidence, a lack of adequate blinding, risk of bias, and the substantial cost of such programs [@wilsonHowEvidencebasedEvidencebased2012].
<!-- Another study Even when negative parenting has been assessed as outcomes, -->
<!-- significant effects were found for positive practices only [@marcynyszynGettingEvidencebasedProgram2011]. -->
<!-- and that there have been prior concerns about the efficacy of such training [@wilsonHowEvidencebasedEvidencebased2012; @cottamDiscoursesUnderpinningParenting2014]. -->
Also note that the efficacy of parenting training has been evaluated primarily in terms of parents' self-reported outcomes;
the ultimate benefit for children,
in particular for adolescents' emotion regulation,
is less studied.
The relatively higher importance rankings for negative versus parenting is in line with the well-established *negativity bias*,
that negative stimuli are more psychologically impactful than positive ones,
which has also been found in children [@vaishNotAllEmotions2008].
Future parenting interventions might take this into consideration by devoting sufficient attention to the elimination of negative parenting practices,
and including children's outcomes, including emotion regulation, into consideration.

@@

The present study had some unique strengths, most notably that it constituted the first comprehensive study of potential predictors of adolescent emotion regulation development. Prior work has taken more piecewise approaches, considering only a few predictors at a time.
<!-- The present study set out to identify the most important early indicators of developmental trajectories -->
<!-- of emotion regulation across adolescence. -->
<!-- This involves three important challenges: -->
<!-- First, the number of potentially relevant risk factors, as identified in a prior text mining systematic review, is large. -->
<!-- Second, there is a paucity of relevant theory to narrow down the candidate predictors. -->
<!-- Third, methods commonly used in the field are not well suited for large numbers of predictors. -->
<!-- Machine learning methods that perform variable selection can overcome these challenges. -->
<!-- We therefore used the SEM-forests algorithm to identify the most important early predictors of trajectories of emotion regulation development throughout adolescence. -->
<!-- The primary aim is to identify the most important predictors of adolescents' trajectories of emotion regulation development. -->
<!-- The secondary aim is to probe the nature of the marginal association of the predictors with these trajectories. -->

The study also has some limitations, most notably that the sample was selective,
which may limit the generalizability of the findings.
Specifically, this was a non-random, ethnically and socially relatively homogenous sample from the Netherlands. The WEIRD-nature of the sample does not allow to generalize this study's findings beyond similar populations.
This raises some specific concerns, for example, that the low importance of SES might be due to a restriction of range.
The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world.
The only way to address these concerns is to replicate these findings in other samples.
Note that all code is available in a reproducible archive, which lowers the threshold to replication.
Another limitation is that the data source was a panel study with mostly questionnaire measures and relatively few biological, cognitive and neurological assays.
One final limitation related to the data is the relatively small sample size.
Machine learning methods benefit from large samples.
Although random forests are relatively robust to overfitting,
larger samples will likely yield more generalizable results,
and also afford more complicated models (thus removing the need for assumptions such as homogeneity of residuals).
Other limitations pertain to the method; for instance, SEM-forests require vast computational resources.
Computational constraints thus prevented us from tuning the model beyond default settings,
and from exploring marginal effects for higher order interactions.
Moreover, at the time of writing, there is not yet a method to evaluate the overall fit of a SEM-forest.
